Practical Parametricity for GADTs

### 2020-07-17 20-gadts/inf-rules.tex
completed proof that obj interps of Nat types are wCPOs

If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then a functor $F :
\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for the
category of $\omega_1$-cocontinuous functors from $\A$ to $\C$, and
$\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
Since (co)limits of functors are computed pointwise, $C^\A$ has all
(co)limits that $\C$ has, and (co)limits of (co)continuous functors
are again (co)continuous. It follows that $[\A,\C]_{(\omega_1)}$ is
($\omega_1$-)(co)complete whenever $\C$ is.  {\color{red} In
  particular, if $F : [\C, \C]_{(\omega_1)}$ for a(n)
  ($\omega_1$-)cocomplete category $\C$, then we can define $F^0 =
  \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at successor ordinals, and
  $F^{\bigcup_{\beta < \alpha} \beta} = \colim{\beta < \alpha}
  F^\beta$ at limit ordinals.}

@@

\new{Need to modify to take equality conditions into account.}  Now,
if $\C$ is any category, then given functors $F,G : \C \to \oCPO$ and
natural transformations $\eta, \eta' : F \to G$, we define $\eta \leq
\eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
natural transformation is clearly the supremum of $(\eta_i)_{i <
  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
\to G\}$ is itself an $\oCPO$.  Letting $F = \lambda \ol
K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
an $\oCPO$.

>>>>

If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then a functor $F :
\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for the
category of $\omega_1$-cocontinuous functors from $\A$ to $\C$, and
$\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
Since (co)limits of functors are computed pointwise, $C^\A$ has all
(co)limits that $\C$ has, and (co)limits of (co)continuous functors
are again (co)continuous. It follows that $[\A,\C]_{(\omega_1)}$ is
($\omega_1$-)(co)complete whenever $\C$ is. In particular, if $F :
[\C, \C]_{(\omega_1)}$ for a(n) ($\omega_1$-)cocomplete category $\C$,
then we can define $F^0 = \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at
successor ordinals, and $F^{\bigcup_{\beta < \alpha} \beta} =
\colim{\beta < \alpha} F^\beta$ at limit ordinals.

@@

%Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
%and natural transformations $\eta, \eta' : F \to G$, we define $\eta
%\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
%$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
%If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
%the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
%x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
%transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
%natural transformation is clearly the supremum of $(\eta_i)_{i <
%  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
%\to G\}$ is itself an $\oCPO$.
%Letting $F = \lambda \ol
%K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
%\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
%that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
%an $\oCPO$.

Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
and natural transformations $\eta, \eta' : F \to G$, we define $\eta
\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
natural transformation is clearly the supremum of $(\eta_i)_{i <  
  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
\to G\}$ is itself an $\oCPO$.  To show that $\oCPOsem{\Gamma;
  \emptyset \vdash \Nat^\Phi \,F\,G}\rho$ is an $\oCPO$, we then still
need to prove that, for any chain of natural transformations
$(\eta_i)_{i < \omega}$ in $\oCPOsem{\Gamma; \emptyset \vdash
  \Nat^\Phi \,F\,G}\rho$, $\bigvee_{i < \omega} \eta_i$ is again in
$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho$.  By
hypothesis, each $\eta_i$ is such that, for all $\ol{K = (K^1, K^2,
  K^*) : \oCPORT_k}$,
\[((\eta_i)_{\overline{K^1}},
(\eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is a morphism in
$\oCPOR$. To see that, for all $\ol{K = (K^1, K^2, K^*) : \oCPORT_k}$,
%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
%%  \omega} \eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
%  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is again a morphism in
%$\oCPOR$, we observe that
$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
  \omega} \eta_i)_{\overline{K^2}})$ is a morphism from
$\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]$ to
$\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi := K}]$ in
$\oCPOR$, we simply observe that
\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i < \omega}
\eta_i)_{\overline{K^2}}) = (\bigvee_{i < \omega}
(\eta_i)_{\overline{K^1}}, \bigvee_{i < \omega}
(\eta_i)_{\overline{K^2}}) = \bigvee_{i < \omega}
((\eta_i)_{\overline{K^1}}, (\eta_i)_{\overline{K^2}})\] so
$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
  \omega} \eta_i)_{\overline{K^2}})$ is the sup of a chain of
morphisms from $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi :=
    K}]$ to $\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi :=
    K}]$ in $\oCPOR$ and thus is itself such a morphism. Taken
together, this all shows that $\oCPOsem{\Gamma; \emptyset \vdash
  \Nat^\Phi \,F\,G}\rho$ is indeed an $\oCPO$.
